---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "./man/figures/README-",
  out.width = "100%"
)
```

# mdgpTutorial

<!-- badges: start -->

[![R-CMD-check](https://github.com/gannd/mdgpTutorial/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/gannd/mdgpTutorial/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The purpose of this tutorial package is to introduce theoretical concepts of scaling and the complexity of scaling categorical data. It then introduces the multi-dimensional grid-point (MDGP) scaling method and the algorithm as implemented in the R package landscapeScaling. Guided exercises demonstrate the effects of user-defined parameters and selection criteria on the definition of scale-appropriate classification schemes that maximize information retention, representativeness of the scaled classes and the detection probability from remotely sensed data. The  MDGP-Scaling algorithm generates new classification schemes on the basis of user desired class label precision of mixed classes and their representativeness across the landscape of interest. Part (1) of the tutorial introduces theory and method to scaling of categorical raster data. Part (2) of the tutorial evaluates the detection of the scaled classes from remotely sensed data.

## Installation

The installation of the tutorial, includes the installation of the newest version of the R package landscapeScaling and its dependencies.
Other required packages for this tutorial are BAMMtools and caret.

```{r setup, include=FALSE}
# install missing packages
packages_installed <- rownames(installed.packages())

# required package names from CRAN
packages <- c('remotes','devtools','terra','sf','leaflet','compositions','caret','BAMMtools')

# install missing packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages],repo = 'https://cran.uni-muenster.de/', dependencies = TRUE)
}

if(!'NLMR' %in% packages_installed){
  remotes::install_github("ropensci/NLMR")
}
# install latest updates of landscapeScaling version
devtools::install_github("gannd/landscapeScaling", force=TRUE)

print('All required packages were installed')

# install.packages("devtools")
devtools::install_github("gannd/mdgpTutorial")
```

